---
title: "`fMRIscrub` Demo"
author: "Isaiah Farrell, Damon Pham, & Amanda Mejia"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{`fMRIscrub` Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}
library(knitr)
knitr::opts_chunk$set(autodep = TRUE, cache = TRUE) # <- warning = FALSE
```

This is the new `fMRIscrub` vignette.

Let's load packages and list the file names:

```{r, warning=FALSE}
library(RNifti)
library(fMRItools)
library(ciftiTools)
library(fMRIscrub)
library(gridExtra)
ciftiTools.setOption("wb_path", "~/Applications/workbench")

rp_fname <- file.path("data", "HCP_Movement_Regressors.txt")
cii_fname <- file.path("data", "HCP_rfMRI_REST2_LR_Atlas_MSMAll.dtseries.nii")
nii_fname <- file.path("data", "HCP_rfMRI_REST2_LR.nii.gz")
```

Cortex plots from `ciftiTools` are made possible by the `rgl` package. To prepare the R Markdown document for knitting we need to do the following: 

```{r}
library(rgl)
rgl::setupKnitr()

# Sometimes the first OpenGL window does not render properly.
rgl::open3d(); rgl::close3d()

# These are also required.
library(manipulateWidget)
library(ggpubr)
```

# CIFTI data

Read:

```{r}
cii <- read_cifti(cii_fname)
```

Plot:

```{r fig.cap="CIFTI cortex data", rgl=TRUE, format="jpg", fig.height=3.5, fig.width=4.5}
view_xifti_surface(cii, title="First volume")
```

```{r eval=FALSE}
view_xifti_volume(cii, title="First volume") # skipped
```

# NIFTI data

Read:

```{r}
# nii <- RNifti::readNifti(nii_fname, volumes = 1:400)
```

Plot:

```{r fig.height=5, fig.width=5}
# RNifti::view(nii[,,,1,drop=FALSE], interactive=FALSE)
```

# RPs

Read:

```{r}
rp <- as.matrix(read.table(rp_fname)[,seq(6)])
dim(rp)
```
With the found rps, we can calculate Framewise Displacement using the `FD` function. By summing the absolute values of the differences in rps, we produce the following results on the time interval:

```{r}
fd <- FD(rp)
FDplot <- plot(fd)
```


# IDvols

Using the `IDvols` function, there are different parameters, namely DVARS and projection scrubbing, you can give for varying outcomes. To start with DVARS, accessed by doing `IDvols(X, method='DVARS')`:

Read:

```{r}
cii_mat <- t(as.matrix(cii))
IDdvars <- IDvols(cii_mat, method='DVARS')
DVARSplot <- plot(IDdvars)
```

The DVARS method used within the `fMRIscrub` package is based on Afyouni and Nichols (2018) dual cutoff method. Similar to FD, DVARS is found by finding the difference in signals. However, FD is based on the motion data, the RPs found earlier. DVARS is based on BOLD signal data, finding the difference between volumes across space.


Plot:

```{r}
grid.arrange(FDplot, DVARSplot, nrow = 2, ncol = 1)
```

The second method mentioned previously was projection scrubbing. Projection scrubbing is used to identify volumes that contain heavy artifacting. To scrub with projection scrubbing, we use `IDvols(X, method = 'projection')`:

Read:

```{r}
TEMP_IDproj_fname <- "Temporary_IDproj.rds"
if (!file.exists(TEMP_IDproj_fname)) {
  IDproj <- IDvols(cii, method = 'projection')
  # IDproj_mat <- IDvols(cii_mat, method = 'projection')
  saveRDS(IDproj, TEMP_IDproj_fname)
}
IDproj <- readRDS(TEMP_IDproj_fname)
```

Plot:

```{r}
XIIplot <- plot(IDproj)
```
